\input{../header.tex}

\newcommand\problemset{4}

\hypersetup{
    pdftitle={physics754 Problem Set \problemset}
}

\newenvironment{aside}{\itshape\small}{}

%\subject{}
\title{physics754 -- Problem Set \problemset}
%\subtitle{}
\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}
\publishers{Group 5 -- Olaf Baake}

\begin{document}

\maketitle

\section{Diffeomorphisms and vector fields}

The various uses of $y$ are not defined in the problem set, so I will do that
now to avoid confusion:

\begin{itemize}
    \item
        $X \colon \R^4 \mapsto \R^4$ is a vector field.
    \item
        $y \colon \R \mapsto \R^4$ is a trajectory.
    \item
        $f_\tau \colon \R^4 \mapsto \R^4$ is a diffeomorphism.
    \item
        $f \colon \R^4 \times \R \mapsto \R^4$ is a $\R$-parameter family of
        diffeomorphisms.
    \item
        $\tilde y \colon \R \times \R^4 \mapsto \R^4$ is a $\R^4$-parameter
        family of trajectories starting at $x$. This is used by $f$ to
        create a $\R$-parameter family of diffeomorphisms.
    \item
        $\hat y \in \R^4$ is a vector in the image of $\tilde y$.
    \item
        $x \in \R^4$ is a vector in the domain of $\tilde y$, which is the
        same as the image, since $y(0) = x$.
    \item
        $\tau \in \R$ is a scalar and parameter for $y$.
\end{itemize}

The ambiguity of round parentheses arising because they are used for both
scoping and function application makes the equation a bit hard to read. I will
use a Mathematica inspired notation that does not contradict the usual
notation: Round parentheses are used only for function application, square
brackets are used for scoping and grouping.

I can now safely omit using different fonts for scalars, vectors and tensors
of higher rank since every symbol is defined.

\subsection{}

With all that set, the first problem nearly does itself:
\begin{align*}
    X(\hat y)
    &= \sbr{\dod{}\tau f_\tau} \del{f_\tau^{-1}(\hat y)} \\
    &= \sbr{\dod{}\tau \tilde y} \del{\tilde y^{-1}(\hat y)} \\
    \intertext{%
        $y$ is a solution of (12) from the problem set. $\tilde y$ additionally
        solves (13) as well. So $\tilde y$ will solve (12). So
        \[
            \sbr{\dod{}\tau \tilde y} (x, \tau) = [X \circ \tilde y](x, \tau)
        \]
        and therefore
    }
    &= [X \circ \tilde y] \del{\tilde y^{-1}(\hat y)} \\
    &= [X \circ \tilde y \circ \tilde y^{-1}] (\hat y) \\
    &= X(\hat y)
\end{align*}

$f_0(x)$ is just $\tilde y(x, 0)$. Since $y(0) = x$, $\tilde y(x, 0)$ reduces
to $x$. Therefore, $f_0$ is indeed the identity transformation. That shows
$f_0 = \mathbb 1$.

\subsection{}

Show that for any $g$:
\[
    \eval{\dod{}\tau f_\tau^* g_{\mu\nu}}_{\tau=0} = 2 X_{(\mu;\nu)},
\]
where I have used the symmetrization parentheses.

The right hand side is given by:
\[
    2 X_{(\mu;\nu)} = 2 X_{(\mu,\nu)} - 2 \Gamma^\lambda_{(\mu\nu)} X_\lambda
\]

Now I will turn to the left hand side, where I will evaluate it at $x$.
\begin{align*}
    \text{RHS}
    &= \eval{\dod{}\tau f_\tau^* g_{\mu\nu}}_{\tau=0} (x) \\
    \intertext{%
        Now I will apply the transformation to $g$.
    }
    &= \eval{\dod{}\tau
    g_{\alpha\beta}\del{f_\tau(x)} \; f_\tau^\alpha{}_{,\mu}(x) \;
    f_\tau^\beta{}_{,\nu}(x)
    }_{\tau=0} \\
    \intertext{%
        The lower index $\tau$ is a little messy, so I will expand $f$ in terms
        of $\tilde y$, introducing even more explicit parameters.
    }
    &= \eval{\dod{}\tau
    g_{\alpha\beta}\del{\tilde y(x, \tau)} \; \tilde y^\alpha_{,\mu}(x, \tau) \;
    \tilde y^\beta_{,\nu}(x, \tau)
    }_{\tau=0} \\
    \intertext{%
        Total differentiation will act on all three factors, since all of them
        depend on $\tau$. This gives three terms:
    }
    &= 
    \sbr{
        \dod{}\tau
        g_{\alpha\beta}\del{\tilde y(x, \tau)}
    }
    \; \tilde y^\alpha_{,\mu}(x, \tau)
    \; \tilde y^\beta_{,\nu}(x, \tau)
    +
    g_{\alpha\beta}\del{\tilde y(x, \tau)}
    \;
    \sbr{
        \dod{}\tau
        \tilde y^\alpha_{,\mu}(x, \tau)
    }
    \; \tilde y^\beta_{,\nu}(x, \tau)
    \\ &\quad
    +
    g_{\alpha\beta}\del{\tilde y(x, \tau)}
    \; \tilde y^\alpha_{,\mu}(x, \tau)
    \;
    \sbr{
        \dod{}\tau
        \tilde y^\beta_{,\nu}(x, \tau)
    }
    \Bigg|_{\tau=0} \\
    \intertext{%
        To simplify, I will now apply the $\tau = 0$ to all the terms that make
        it possible at this stage. $\tilde y(x, 0)$ is just $y(0)$, which was
        defined to be $x$. This simplification will be in the $g(\ldots)$. The
        other thing I can do right now is to look at $\tilde y^\alpha_{,\mu}(x,
        \tau)$: When $\tau = 0$, $\tilde y$ reduces to $x$, so that will be a
        $\delta^\alpha_\mu$.
    }
    &=
    \sbr{
        \dod{}\tau
        g_{\alpha\beta}\del{\tilde y(x, \tau)}
    } \Bigg|_{\tau=0}
    \; \delta^\alpha_\mu
    \; \delta^\beta_\nu
    +
    g_{\alpha\beta}(x)
    \;
    \sbr{
        \dod{}\tau
        \tilde y^\alpha_{,\mu}(x, \tau)
    } \Bigg|_{\tau=0}
    \; \delta^\beta_\nu
    \\ &\quad
    +
    g_{\alpha\beta}(x)
    \; \delta^\alpha_\mu
    \;
    \sbr{
        \dod{}\tau
        \tilde y^\beta_{,\nu}(x, \tau)
    } \Bigg|_{\tau=0}
    \\
    \intertext{%
        The $\tau$-differentiation for $\tilde y$ are given via the
        differential equation that defines $y$.
        \[
            \dod{}\tau \tilde y^\alpha_{,\mu}(x, \tau)
        \]
        can be written as
        \[
            \dod{}\tau
            \dpd{}{{x^\mu}}
            \tilde y^\alpha(x, \tau).
        \]
        The derivatives should commute, so this becomes
        \[
            \dpd{}{{x^\mu}}
            \dod{}\tau
            \tilde y^\alpha(x, \tau)
            =
            \dpd{}{{x^\mu}}
            X^\alpha\del{y(\tau)}
        \]
        via the differential equation (12). There is no problem with setting
        $\tau = 0$ here. $y(\tau)$ becomes $x$. I will write the result as
        $X^\alpha_{,\mu}(x)$.
    }
    &=
    \eval{
        \dod{}\tau
        g_{\mu\nu}\del{\tilde y(x, \tau)}
    }_{\tau=0}
    +
    g_{\alpha\nu}(x)
    \;
    X^\alpha_{,\mu}(x)
    +
    g_{\mu\beta}(x)
    \;
    X^\beta_{,\nu}(x)
    \\
    &= \eval{
        \dod{}\tau
        g_{\mu\nu}\del{\tilde y(x, \tau)}
    }_{\tau=0}
    + 2 X_{(\mu,\nu)}(x) \\
    \intertext{%
        Now comes the part where the $\Gamma$ have to appear to make it fit.
        Let $\hat y := \tilde y(x, \tau)$ from this point. That lets me write
        the chain rule:
    }
    &= \eval{
        \frac{\partial g_{\mu\nu}}{\partial \hat y^\lambda}
        \dpd{{{\hat y}^\lambda}}\tau
    }_{\tau=0} (x)
    + 2 X_{(\mu,\nu)}(x) \\
    \intertext{%
        The $\tau$-derivative of $\hat y$ can be expressed by the differential
        equation. $g$ will be differentiated with respect to $x^\lambda$ in the
        $\tau = 0$ case.
    }
    &=
    g_{\mu\nu,\lambda}(x)
    \; X^\lambda(x)
    + 2 X_{(\mu,\nu)}(x) \\
    &=
    \sbr{
        g_{\mu\nu,\lambda}
        \; X^\lambda
        + 2 X_{(\mu,\nu)}
    }(x)
\end{align*}

The $\Gamma$ terms are missing, since there is only derivative of $g$.

\subsection{Action}

Given
\[
    S_\text{gr}(g) := - \frac{1}{16\piup G} \intop_{\R^4} \dif{^4 x} \,
    \sqrt{|g|} \mathcal R(g)
\]
I have to show that
\[
    \eval{\dod{}\tau S_\text{gr}(f^*_\tau g)}_{\tau=0} = \frac{2}{16\piup G}
    \intop_{\R^4} \dif{^4 x} \, \sqrt{|g|} G^{\mu\nu} X_{(\mu;\nu)}.
\]

The various forks induced by the product rule will make this calculation
nonlinear, I will try to perform an in-order traversal of all the steps while
dragging intermediate results along. First I will put the transformed $g$ into
the definition of the action.
\[
    \eval{\dod{}\tau S_\text{gr}(f^*_\tau g)}_{\tau=0}
    = - \frac{1}{16\piup G} \eval{\dod{}\tau \intop_{\R^4} \dif{^4 x} \,
    \sqrt{|f^*_\tau g|} \mathcal R(f^*_\tau g)}_{\tau=0}
\]
Given that the result has a similar prefactor and an integration, I will focus
my attention to the integrand. If some summand will be left that does not fit
into the result, I will see whether the integral over it does vanish. First I
will assume that I can interchange the limit, the differentiation and the
integration.
\begin{align*}
    &
    \eval{\dod{}\tau \sqrt{|f^*_\tau g|} \mathcal R(f^*_\tau g)}_{\tau=0} \\
    &= 
    \eval{\sbr{\dod{}\tau \sqrt{|f^*_\tau g|}} \mathcal R(f^*_\tau g)}_{\tau=0}
    +
    \eval{\sqrt{|f^*_\tau g|} \dod{}\tau \mathcal R(f^*_\tau g)}_{\tau=0}
    \\
    \intertext{%
        The terms that are not differentiated to not have to wait until $\tau$
        is set to zero. Once this is done $f$ will become the identity
        transformation, leaving the $g$ untouched.
    }
    &=
    \sbr{\dod{}\tau \sqrt{|f^*_\tau g|}}_{\tau=0} \mathcal R(g)
    + \sqrt{|g|} \sbr{\dod{}\tau \mathcal R(f^*_\tau g)}_{\tau=0}
    \intertext{%
        Using Jacobi's formula for the derivative of determinants that we
        proved on an earlier exercise set I will work on the first term in
        square brackets.
    }
    &=
    \sbr{\frac1{\sqrt{|f^*_\tau g|}} \Tr\del{[f^*_\tau g]^\dagger \dod{}\tau
    f^*_\tau g}}_{\tau=0} \mathcal R(g)
    + \sqrt{|g|} \sbr{\dod{}\tau \mathcal R(f^*_\tau g)}_{\tau=0}
    \intertext{%
        The first part can be evaluated at $\tau = 0$ now. The last derivative
        was derived in the prior parts, so that I just plug that in.
    }
    &=
    \frac1{\sqrt{|g|}} \Tr\del{g^\dagger 2 X_{(\mu;\nu)}} \mathcal R(g)
    + \sqrt{|g|} \sbr{\dod{}\tau \mathcal R(f^*_\tau g)}_{\tau=0}
    \intertext{%
        That trace will totally contract the $g$ and the $X$, giving a scalar:
    }
    &= \frac2{\sqrt{|g|}} g^{\mu\nu} \mathcal R(g) X_{(\mu;\nu)}
    + \sqrt{|g|} \sbr{\dod{}\tau \mathcal R(f^*_\tau g)}_{\tau=0}
    \intertext{%
        I swapped the $g$ and the $X$ in the first summand, since I want to
        factor out $X$ later on since $g \mathcal R$ appear in the Einstein
        tensor which gets contracted with $X$ in the final expression. The next
        work will be done on the second summand. First, I expand the Ricci
        scalar in terms of the Ricci tensor.
    }
    &= \frac2{\sqrt{|g|}} g^{\mu\nu} \mathcal R(g) X_{(\mu;\nu)}
    + \sqrt{|g|} \sbr{\dod{}\tau \sbr{f^*_\tau g^{\mu\nu}} R_{\mu\nu}(f^*_\tau g)}_{\tau=0}
    \intertext{%
        While I apply product rule, I will evaluate at $\tau = 0$ in this step
        as well. The ellipsis stands for the first summand of the above
        expression.
    }
    &= \ldots
    + \sqrt{|g|} \sbr{\dod{}\tau f^*_\tau g^{\mu\nu}}_{\tau=0} R_{\mu\nu}(g)
    + \sqrt{|g|} g^{\mu\nu} \sbr{\dod{}\tau R_{\mu\nu}(f^*_\tau g)}_{\tau=0}
    \intertext{%
        The dual metric tensor is defined as $g^{\alpha\beta} g_{\beta\gamma} =
        \delta^\alpha_\gamma$. This also holds for the transformed metric
        tensor. \begin{aside}The $\delta$ is not transformed since it is a
        symbol and not a tensor?\end{aside} Transforming that identity yields
        \[
            \sbr{f^*_\tau g^{\alpha\beta}} \sbr{f^*_\tau g_{\beta\gamma}} =
            \delta^\alpha_\gamma.
        \]
        The derivative with respect to $\tau$ gives, applying the product rule
        right away and evaluating at $\tau = 0$:
        \[
            \sbr{\dod{}\tau f^*_\tau g^{\alpha\beta}}_{\tau = 0}
            g_\beta\gamma
            +
            g^{\alpha\beta}
            \sbr{\dod{}\tau f^*_\tau g_{\beta\gamma}}_{\tau = 0}
            = 0.
        \]
        Using the known expression for the derivative in the second summand,
        this can be written as
        \[
            \sbr{\dod{}\tau f^*_\tau g^{\alpha\eta}}_{\tau = 0}
            = - 2 X^{(\alpha;\eta)}.
        \]
        The $^{;\eta}$ includes the metric tensor. Now this goes back into the
        dragged along expression. Since the line is long enough, I included the
        omitted part again. I also raised and lowered the indices on $R$ and
        $X$ respectively, to remove the $^{;\eta}$ that was not featured in the
        lecture so far.
    }
    &= \frac2{\sqrt{|g|}} g^{\mu\nu} \mathcal R(g) X_{(\mu;\nu)}
    -2 \sqrt{|g|} R^{\mu\nu}(g) X_{(\mu;\nu)}
    + \sqrt{|g|} g^{\mu\nu} \sbr{\dod{}\tau R_{\mu\nu}(f^*_\tau g)}_{\tau=0}
    \intertext{%
        The middle summand is the part with $R$ within $G$. It has the correct
        sign and prefactor. Now the first and last term have to turn out to be
        $\sqrt{|g|} \mathcal R g X$. The factor $\frac12$ that is within $G$
        will be supplied by the symmetrization of $X$.
    }
\end{align*}

\IfFileExists{\bibliographyfile}{
    \printbibliography
}{}

\end{document}

% vim: spell spelllang=en tw=79
